Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-10-04 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

Thank you for your helpful comments. We think that this version of the paper is significantly improved, and we hope you agree.

Outline of major revisions:
1. We have generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal. The previous version presented an argument that the ML branch lengths would have this property, but this argument did not hold up to further scrutiny.
2. We have expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation.
3. We have changed discussion and plots to involve branch lengths and not fidelities.

@@

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

Thank you for the opportunity to improve on our analytically-derived regions of inconsistency.
Indeed, obtaining these regions in closed form is feasible, though it involves comparing a large number of two-variable functions to one another.
Fig. S2 shows the regions where a particular likelihood equation will be maximal, and allows us to compute---in closed form---the value of the branch parameter in Fig. 2.

@@

> Reviewer: 2

> Comments to the Author
> In this manuscript  the authors show that joint inference of phylogeny
> and ancestral characters is not consistent in general by providing a
> counter-example, in which as the number of sites grows, they show that
> the wrong tree topology can have a higher likelihood than the correct
> topology under the joint inference model.
> The result is important for the community to understand, in particular
> because TreeTime (the joint inference method) is used in high-profile
> analyses on Ebola, influenza and other viruses on the nextflu /
> nextstrain websites. These are beautiful and accessible sites aiming
> to make analyses on viruses open to exploration by the community, but
> this comes at the cost of suppressing the details of the analysis. If
> indeed there are deep statistical issues or limitations it is
> important that the community knows what these are and what their
> implications are.

> Accordingly my comments are focussed on where I think the manuscript
> could be more clear (and where I did not quite follow the logic). My
> major suggestion would be to go even further towards demonstrating in
> a simple example how the joint inference can be wrong – for example,
> providing a toy multiple sequence alignment or even one of some
> influenza HA sequences – you could find 2 sets of 4 closely-related
> sequences that have the Farris and Felsenstein topologies, maybe? And
> demonstrate that the inference gives the wrong result by explicitly
> computing the likelihood.

We agree that understanding the implications of our derivation is a very interesting avenue for future work.
However, to do such a project justice would require a separate paper.
We suspect that once our paper demonstrates this inconsistency, others will quickly follow to write such a paper.

@@

> Also, laying out the theory a little bit more step-by-step so that a
> general quantitative reader doesn't struggle too much to follow would
> help (suggestions below).

Thank you very much for reading the derivations and for your suggestions.
As noted inline we have revised the text accordingly.

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

We have considered how to extend these results to larger trees but this does not appear to be straightforward.

@@

> L 132 clarify “equivalent but different notion” – different from what?

We have reformulated this sentence to avoid the comparison:
"We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called "fidelity.""

> - line 146 “seeing no change along a branch more likely increases the
> likelihood by introducing a term of (1 + θ) instead of (1 − θ)” – I
> don't understand this

We have clarified the intuition in these new sentences:
"For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth."

@@

> Reviewer: 3

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but as we show the bias in branch lengths persists even in these cases.

We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
"Although \citet{Sagulenko2018-xl} explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference."

@@

> Comments to the Author
> Review of Shaw, D. and Matsen, F.  "Joint maximum-likelihood of
> phylogenies and ancestral states is not consistent"

> Shaw and Matsen demonstrate that the approach of jointly estimating
> ancestral states and tree topologies (or branch lengths on a fixed
> topology) is not statistically consistent.  I believe that their
> results are correct, and that their result is important to
> disseminate.  The paper is clearly written, given the complexity of
> the material, and I have no reservation about recommending publication
> virtually as is.

> However, although my original intent was to carefully work through the
> math (including the derivations and proofs in the appendix), I have
> not found the time to do so, and because of the lateness of my review
> at this point, I will provide just a few comments in the hope that
> they will be useful.  Hopefully, other reviewers were able to check
> the correctness of the mathematical results more carefully.

Thank you for your assessment.
We have worked hard in our initial submission and in our revision to write out the argument in great (indeed, excruciating) detail, and hope that you will be able to read the revision more fully.

### 2018-10-05 appendix.tex
Pass through.

For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.

@@

In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.

@@

\begin{proof}
For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the log-likelihood proportional to
\[
\ell(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
\[
h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

@@

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

We show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$.
It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
The same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
\end{proof}

Examples of generating parameters resulting in consistency of joint inference occur in the following.
Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
\]
There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
In this case, $\hat{x}_1 = x^*$ implies consistency only occurs when $x^*=0$ or $x^*=1$.
Another case,
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
\]
shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}

The boundaries determining maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are obtained through maximizing over 81 separate functional values (Table~\ref{tab:likelihoods}).
Using the notation in the proof of Theorem~\ref{thm:bl_theorem}, we see that, in general, our likelihood given in~\eqref{eq:site_pattern_profile_likelihood_mean} has an exact, closed form maximum of
\begin{align}
\ell_{\tau^*,t^*}(\tau_1, \hat{t}; \boldsymbol\xi) &= \frac{1+\hat{x}_1}{2}\log(1+\hat{x}_1) + \frac{1-\hat{x}_1}{2}\log(1-\hat{x}_1)\nonumber\\
&\qquad + \frac{1+\hat{y}_1}{2}\log(1+\hat{y}_1) + \frac{1-\hat{y}_1}{2}\log(1-\hat{y}_1)\nonumber\\
&\qquad + \frac{1+\hat{x}_2}{2}\log(1+\hat{x}_2) + \frac{1-\hat{x}_2}{2}\log(1-\hat{x}_2)\nonumber\\
&\qquad + \frac{1+\hat{y}_2}{2}\log(1+\hat{y}_2) + \frac{1-\hat{y}_2}{2}\log(1-\hat{y}_2)\nonumber\\
&\qquad + \frac{1+\hat{w}}{2}\log(1+\hat{w}) + \frac{1-\hat{w}}{2}\log(1-\hat{w}).
\end{align}
For example, in the lower right region of Fig.~\ref{fig:max-anc-state}, the maximal ancestral state split is $\hat{\boldsymbol\xi}_1 = \{\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset,\emptyset\}$ so that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \{x^*(y^*)^2,(y^*)^2,x^*(y^*)^2,1,1\}
\]
and the exact, closed form value of the likelihood is
\begin{align}
\ell_{U,1}(x^*,y^*) &= (1+x^*(y^*)^2)\log(1+x^*(y^*)^2) + (1-x^*(y^*)^2)\log(1-x^*(y^*)^2)\nonumber\\
&\qquad + \frac{1+(y^*)^2}{2}\log(1+(y^*)^2) + \frac{1-(y^*)^2}{2}\log(1-(y^*)^2)\nonumber\\
&\qquad + 2\log(2).
\end{align}
We obtain similar maximal values as functions of $x^*$ and $y^*$ for each ancestral state split.
The curves delineating maximal ancestral state splits (Fig.~\ref{fig:max-anc-state}) are determined by these 81 functions.

>>>>

%EM This is very minor compared to the below comments, so address it second in a separate commit, please, if you wish to address it. Curly braces are really for unordered sets, right? I suggest round parensI suggest round parens.
For a fixed topology $\tau$, we define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization $\ancestralStateColumn_i$.
In this case we cannot use the same complement trick as before: the probability of observing an ancestral state split conditional on a site split is not invariant to taking its complement.
We thus define an ``ancestral state split'' $\ancestralSplit$ for an internal node $\ancestralStateColumn_i$ to be the set of internal nodes labeled with $1$ if the $\nSiteRows$th tip is not labeled with $1$, and as the set of internal nodes labeled with $1$ in $\overline{\ancestralStateColumn}_i$ if the $\nSiteRows$th tip is labeled with $1$.
We emphasize that the ancestral state split complementing procedure depends on tip states, not ancestral states: both site splits and ancestral state splits are defined by whether the $\nSiteRows$th element of $\alignmentColumn_i$ is labeled as $1$.

@@

In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node states.

@@

\begin{proof}
For a fixed, known $\boldsymbol\xi$, there exists a closed form solution to $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ solving
\[
\hat{t}_{\boldsymbol\xi} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi).
\]
We show in this case that the log-likelihood $\ell$ attains a unique maximum at $\hat{t}_{\boldsymbol\xi}$.
For fixed $\boldsymbol\xi$, the log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
%EM These c_j's and their properties need to be explained a lot more clearly. What are they, how can we know that the likelihood has this form, and why they have the properties below.
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t, \boldsymbol\xi) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the log-likelihood proportional to
\[
\ell(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
\[
h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
%EM Here's this propto thing again. It's up to a constant, not propto, right?
\ell(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1), \label{eq:closed-form-x}
\end{align}
which has second derivative
\[
\ell''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1]$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
%EM We need to hear why both of these properties hold.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\ell''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero, and $\hat{t}$ uniquely maximizes $\ell$.

@@

%EM Isn't this table redundant given Table S3?
\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_{\siteSplit_j}$ & $32\cdot\Pr(\siteSplitRV=\siteSplit_j,\ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ where $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

We show that solutions of this form never obtain $\hat{t} = t^*$ except on a set of measure zero.
Given Table~\ref{tab:gen-sitepatprob}, all solutions $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$.
%EM The following sentence falls short of getting us rigorously to where we need to go, which is the measure zero statement. Doesn't that just follow from the number of zeros of a polynomial being finite?
It is possible to obtain $\hat{x}_1=x^*$, such as when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch lengths is zero or infinite, but we cannot have $\hat{x}_1=x^*$ for all $x^*$.
The same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent on a set of measure zero.
\end{proof}

%EM I don't see this paragraph being helpful.
%EM It's a confusing read. First you say that we're going to look at examples, but then perhaps you cast doubt on if the examples exist "There may be no..."
Examples of generating parameters resulting in consistency of joint inference occur in the following.
Let $x^*$ and $y^*$ be such that the estimate for $\hat{x}_1$ is
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{123} + p_{13}\right) - 1 = (x^*)^2.
\]
There may be no $x^*$ and $y^*$ that yield this maximum (Fig.~\ref{fig:max-anc-state}), though it is a feasible solution to the general joint inference problem.
In this case, $\hat{x}_1 = x^*$ implies consistency only occurs when $x^*=0$ or $x^*=1$.
Another case,
\[
\hat{x}_1 = 2 \cdot \left(p_{\emptyset} + p_2 + p_{3} + p_{123} + p_{23} + p_{13}\right) - 1 = \frac{1}{2} + \frac{1}{2}(x^*)^2
\]
shows $\hat{x}_1 = x^*$ implies consistency only when $x^*=1$.

@@

The regions in Fig.~\ref{fig:max-anc-state} are analytically-derived regions of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
As the region of degeneracy in Fig.~\ref{fig:max-anc-state} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
It is easy to see that when $\emptyset$ is the maximal ancestral state split, we have the same log-likelihood for $\tau_1$ and $\tau_2$.
Moreover, if $w=1$, the internal branch becomes zero-length and the two topologies are indistinguishable.
%EM The rest of this paragraph is obvious, no?
Let $\mathcal{T}_0$ be such that, for $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, $t^*\in\mathcal{T}_0$ corresponds to $x^*$ and $y^*$ falling in the region in Fig.~\ref{fig:max-anc-state} where $\hat{\boldsymbol\xi}=\hat{\boldsymbol\xi}_1$.
We can see this results in the likelihood of both topologies being equal, i.e.,
\begin{align*}
\max_{t : t^*\in\mathcal{T}_0} & \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_1} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\})\\
&= \max_{t: \boldsymbol\xi=\hat{\boldsymbol\xi}_1, w=1, \tau=\tau_2} \ \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_{\siteSplit_j} \mid \tau, \{x_1,y_1,x_2,y_2,w\}).
\end{align*}
